```{r}
library(dplyr)
library(ggplot2)
library(patchwork)
library(treemapify)
library(reshape2)
library(plyr)
```


```{r}
usa_deaths_states = read.csv("data/death-causes-usa.csv", sep=";")
spain_deaths = read.csv("data/death-causes-spain-2017-modified.csv", sep=";",fileEncoding="UTF-8-BOM")
```



USA dataset


The data was published by the Centers for Disease Control and Prevention, gathered by the National Center for Health Statistics (NCHS) and last updated on 2017. This dataset presents the age-adjusted death rates for the 10 leading causes of death in the United States. The data is based on information from all resident death certificates filed in the 50 states and the District of Columbia using demographic and medical characteristics. Age-adjusted death rates (per 100,000 population) are based on the 2000 U.S. standard population.

The information we will use in this dataset leaves us with 10868 observations and 6 columns (Year, Technical Cause Name, Cause Name, State, Deaths and Age-adjusted Death Rate).

```{r}
dim(usa_deaths_states)
head(usa_deaths_states)
```

Spanish dataset


Spanish data was found through a goverment webpage (datos.gob.es), and is part of a study done by the Spanish Institute of Statistics (INE), for the year 2017. We adapted the death causes in Spanish dataset to match the ones in the US one, as the Spanish dataset was more detailed (more details in the file disease-comparison.xlsx). After doing this, we see that the dataset has 1056 observations and 4 columns (Disease, Gender, Age and Number of Deaths).


```{r}
dim(spain_deaths)
head(spain_deaths)
```


As we see, we have different data for both countries, so the comparison will be tough. We will first filter for the country data, filtering state data in the US dataset in order to view the country as a whole.

```{r}
usa_deaths = usa_deaths_states %>% filter(State == "United States") %>% filter(Cause.Name != "All causes")
head(usa_deaths)
```


```{r}
spain_deaths %>% filter(GENDER == "Both") %>% filter(AGE=="All ages") %>% View()
```


```{r}
ggplot(usa_deaths, aes(x=Year, y=Age.adjusted.Death.Rate, fill=Cause.Name)) + 
  labs(title = "Trend of death causes", x = "", y = "Death Rate / 100.000 (Age Adjusted)", fill = "Causes") +
  scale_color_brewer(palette = "Paired") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme_minimal() +
  geom_area()

ggplot(usa_deaths, aes(x=Year, y=log(Deaths), color=Cause.Name)) + 
  labs(title = "Changes on death causes over the years", x = "", y = "log(Death Rate)", fill = "Causes") +
  scale_color_brewer(palette = "Paired") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme_minimal() +
  geom_line()
```



```{r}

#Get data about the top 10 diseases in the Spanish data set. Ignore "Other causes"
topTenSpanishDiseases<- spain_deaths %>% filter(GENDER == "Both") %>% filter(AGE=="All ages") %>% filter(DISEASE!="All causes") %>% filter(DISEASE!="Other causes") %>% top_n(10, NUMBER.OF.DEATHS)

#Generate tree map
ggplot(topTenSpanishDiseases, aes(area=NUMBER.OF.DEATHS, fill=DISEASE, label=NUMBER.OF.DEATHS)) +
  scale_fill_brewer(palette = "Paired") +
  labs(title = "Most deadly diseases in Spain (2017)", fill = "Causes") +
  geom_treemap() +
  geom_treemap_text(fontface = "italic",
                    colour = "white",
                    place = "centre",
                    grow = FALSE,
                    reflow = TRUE) 
```

```{r}

#Create data frame to hold data about people who have Diseases of the circulatory system
diseasesCircSystemData <- spain_deaths

#Set up AGE factor for pyramid plot
diseasesCircSystemData$AGE <- factor(diseasesCircSystemData$AGE, c("0 to 1", "1 to 4", "5 to 9", "10 to 14", "15 to 19", "20 to 24", "25 to 29", "30 to 34", "35 to 39", "40 to 44", "45 to 49", "50 to 54", "55 to 59", "60 to 64", "65 to 69", "70 to 74", "75 to 79", "80 to 84", "85 to 89", "90 to 94", "95 or more", "All ages"))

#Filter data to required 
diseasesCircSystemData <- diseasesCircSystemData %>% filter(DISEASE == "Diseases of the circulatory system") %>% filter(GENDER != "Both") %>% filter(AGE != "All ages") 

#Mutate Male data so it will be negative on bargraph
diseasesCircSystemData <- diseasesCircSystemData %>% mutate(NUMBER.OF.DEATHS = ifelse(GENDER == "Males", -1 * NUMBER.OF.DEATHS, NUMBER.OF.DEATHS))

#Generate pyramid plot
 ggplot(diseasesCircSystemData, aes(x = AGE, y = NUMBER.OF.DEATHS, fill = GENDER)) + 
  geom_bar(data=diseasesCircSystemData[diseasesCircSystemData$GENDER == "Females",], stat = "identity") + 
  geom_bar(data=diseasesCircSystemData[diseasesCircSystemData$GENDER == "Males",], stat = "identity") + 
  scale_y_continuous(breaks = seq(-15000, 15000, 5000), 
                     labels = paste0(as.character(c(seq(15, 0, -5), seq(5, 15, 5))), "")) +
  coord_flip(ylim=c(-15000,15000)) + 
  scale_fill_brewer(palette = "Set1") + 
  labs(title = "Number of people who die from Diseases of the circulatory system", y = "Number of deaths (1,000)", x = "Age", fill = "Gender") +
  theme_bw()
```


```{r}
```



